---
title: "Random Forest vs ARIMA for Stress State Prediction"
subtitle: "This is a Report Template Quarto"
author: "Indupriya Uppaluri (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

Nice report!
:::

## Introduction

### Paper 1: Stress Detection by Machine Learning and Wearable Sensors

Garg et al. (2021) investigated the use of machine learning models to detect stress using the WESAD dataset, which contains physiological signals such as ECG, body temperature, respiration, EMG, and electrodermal activity from wearable sensors. The study compared several models including k-NN, Linear Discriminant Analysis, Random Forest, AdaBoost, and Support Vector Machine for both binary (stress vs. non-stress) and three-class (neutral vs. stress vs. amusement) classification. Random Forest outperformed all other models, achieving an F1-score of 83.34 for binary classification and 65.73 for three-class classification. This paper is directly relevant to our project as it demonstrates Random Forest's effectiveness on the exact dataset we are using.

### Paper 2: Introducing WESAD, a Multimodal Dataset for Wearable Stress and Affect Detection

Schmidt et al. (2018) introduced WESAD, a publicly available multimodal dataset designed for wearable stress and affect detection. The dataset was collected from 15 subjects using both wrist-worn and chest-worn devices, capturing physiological signals including electrocardiogram (ECG), electrodermal activity (EDA), electromyogram (EMG), respiration, body temperature, blood volume pulse, and three-axis acceleration. The study measured three affective states: neutral (baseline), stress, and amusement. Results showed that using chest-based physiological modalities alone achieved strong classification performance, with 79% accuracy for three-class and 92% for binary classification. This paper is foundational to our project as it describes the exact dataset we are using and establishes the key physiological signals most relevant to stress detection.

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{python}
import pandas as pd
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
