---
title: "Stress State Prediction from Wearable Sensors: A Comparative Study of ARIMA and Random Forest on the WESAD Dataset"

author: "Indupriya Uppaluri, Virginia Vaughan, Chawan Sreelekha (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)



## Introduction


Stress detection in real-time, and more importantly biomarkers, have gained interest as both military and corporate facilities aim to improve the quality of worker health and performance. This has let to extensive research using commercial wearables that record signals such as beats per minute (BPM), heartrate (HR), heart rate variability (HRV), galvanic skin conductance (EDA), respiration (RESP), and temperature (TEMP). These physiological signals are used in tandem with both stochastic models such as ARIMA and machine learning such as Random Forest. Both types of models offer their own positives and negatives when predicting physiological data. For example, ARIMA is a more traditional, linear-based model, requires less data for training and extrapolation, while Random Forest is more data-intensive but does not assume linearity, making it better suited for modeling complex patterns. 

In this paper, we hope to explore whether it is possible for the simpler ARIMA model to predict stress state at the same accuracy of a machine learning model like Random Forest. If the ARIMA can perform similar to or greater, this makes an argument that it is still possible to predict stress in physiological signals without the intensive training needed in machine learning, despite ARIMA’s assumptions and limitations. To explore this, we chose the WESAD dataset, which contains multimodal physiological data collected from participants as they underwent different activities, such as no-stress, stress, and amusement. Based on our literature review, we have found that the best physiological signals for predicting stress state that are also provided by the dataset include HR, HRV, and EDA, and that these metrics can be enhanced with secondary metrics such as RESP and TEMP. One paper, [@oyeleye2022] found that the accuracy of ARIMA prediction can be heavily improved using the “rolling window” approach. This helped the model predict based on the signal during different windows of time instead of treating each sample independently of the other. In our methodology we plan to create a pipeline to preprocess the data, compute our feature metrics, derive rolling windows, and predict a given metric using both ARIMA and Random Forest in order to compare the model’s accuracy. 


## Literature review


ARIMA, which stands for Autoregressive Integrated Moving Average, is a classical statistical model widely used for time-series forecasting that captures temporal dependencies within a signal using its own past values and error terms. [@kontopoulou2023] found that ARIMA remains a broadly used and interpretable approach, particularly well-suited for structured, lower-dimensional datasets where its linearity assumptions hold reasonably well. Critically, [@ziyadidegan2025] identified ARIMA as an underutilized but promising method specifically for physiological stress research, appearing in only 3 of 119 reviewed cardiovascular stress studies despite its recognized capacity to capture temporal rhythms in cardiovascular and electrodermal signals. Given that ECG-derived metrics reflect the electrical activity of the heart and EDA captures the skin's electrodermal response to psychological arousal, both of which change measurably during stress, the temporal structure of these signals makes them theoretically well-suited for time-series modeling through ARIMA with a rolling window approach. If stress-related changes in ECG and EDA follow predictable temporal patterns, ARIMA may be capable of detecting stress states without the computational demands of machine learning, a question this study directly investigates. 


Random Forest is an ensemble machine learning method that builds multiple decision trees and aggregates their outputs to produce robust classifications, making no assumptions about the linearity or distribution of the underlying data. [@garg2021] evaluated Random Forest alongside several other classifiers including k-NN, Linear Discriminant Analysis, AdaBoost, and Support Vector Machine on the WESAD dataset, finding that Random Forest achieved consistently strong performance across both binary stress versus non-stress and three-class neutral, stress, and amusement classification tasks. [@schmidt2018], who originally introduced the WESAD dataset, a multimodal physiological signal collection from 15 subjects recorded across controlled stress, baseline, and amusement conditions using wrist-worn and chest-worn devices capturing signals including ECG and EDA among others, also conducted preliminary Random Forest experiments achieving 88.33% baseline accuracy, further establishing it as the benchmark comparator for this study. Together, these findings position Random Forest as a well-validated and high-performing method for stress classification on WESAD, and the central question of this study is whether the simpler, interpretable ARIMA model can achieve comparable predictive performance on ECG and EDA signals alone.


## Methods

### Random Forest Classification Framework

The predictive model used for multi-class affective state classification is the non-parametric Random Forest algorithm. Formally, a Random Forest is an ensemble classifier consisting of a collection of structured decision trees $\{h(x, \theta_k), k = 1, \dots\}$ where the parameters $\{\theta_k\}$ are independent, identically distributed random vectors, and each individual tree casts a single unit vote for the most popular class at input feature vector $x$.

The foundational non-parametric regression and classification framework can be defined by the following sequence:

$$Y_i = m(X_i) + \varepsilon_i$$

where $Y_i$ represents the target affective stress label ($Y_i \in \{1, 2, 3\}$: Neutral, Stress, Amusement), which is defined as the sum of the true underlying physiological mapping function value $m(x)$ for the feature tensor $X_i$, and $\varepsilon_i$ represents the random observation errors.

In this implementation, the target function $m(x)$ is mathematically unknown. With the help of this definition, we create a robust local averaging estimation by aggregating a multitude of decorrelated randomized decision trees ($T$). The ultimate classification function estimation formula is printed below:

$$\hat{m}(x) = \text{mode} \left\{ T_1(x), T_2(x), \dots, T_B(x) \right\}$$

In other words, this means that we are discovering the emotional state boundaries across the physiological multi-sensor space with the help of bootstrap aggregation (bagging) and randomized feature sub-selection across $B$ independent trees ($B = 100$). This collective voting structure decreases the overall structural variance of individual decision trees without increasing the model's bias.

#### Justification of Choices Based on Problem and Data
We chose the Random Forest algorithm for this comparative time-series framework based on specific properties of the WESAD wearable dataset:

- **Handling of Non-Linear Physiological Interaction:** Autonomic nervous system responses do not follow strict linear bounds; a spike in skin conductance (EDA) combined with a drop in skin temperature (TEMP) creates highly complex, non-linear patterns during stress episodes.
- **Robustness to Sensor Artifacts and Noise:** Wearable sensor data collected in the field is prone to sudden movement artifacts. Because Random Forest splits nodes by evaluating localized subsets of data and features via the Gini impurity index, it prevents individual extreme outliers or sensor errors from corrupting the global classification boundaries.

### Implementation Script
*The common non-parametric regression model is* $Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is unknown and* $\varepsilon_i$ *some errors. With the help of this definition, we can create the estimation for local averaging i.e.* $m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$ *is near to* $x$*. In other words, this means that we are discovering the line through the data points with the help of surrounding data points. The estimation formula is printed below [@R-base]:*

$$ M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i \tag{1} $$

*$W_n(x)$ is the sum of weights that belongs to all real numbers. Weights are positive numbers and small if $X_i$ is far from $x$.*

#### Justification of Choices Based on Problem and Data
This non-parametric local averaging framework (Equation 1) is explicitly chosen for our WESAD wearable sensor prediction task instead of standard parametric regression algorithms, which typically assume a strict linear profile [WESAD]:

$$ y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i \tag{2} $$

A parametric linear model (Equation 2) assumes a constant, straight-line relationship that completely fails when applied to the complex, non-linear biological response profiles found within multi-modal wearable sensor streams [WESAD]. We chose the Random Forest framework based on three specific data properties that align with this non-parametric local estimation approach:

- **Handling of Non-Linear Physiological Interaction:** Autonomic nervous system responses do not follow strict linear bounds; a spike in skin conductance (EDA) combined with a drop in skin temperature (TEMP) creates highly complex, non-linear patterns during stress episodes [WESAD]. Random Forest handles these multi-modal, non-linear relationships out of the box through step-wise feature space partitioning without requiring restrictive linear coefficients like those in Equation 2.
- **Robustness to Sensor Artifacts and Noise:** Wearable sensor data collected in the field is prone to sudden movement artifacts and temporary data dropouts [WESAD]. Because Random Forest splits nodes by evaluating localized subsets of data and features via the Gini impurity index, it acts as a robust local estimator, preventing individual extreme outliers or sensor errors from corrupting the global classification boundaries.
- **Resistance to Temporal Overfitting:** High-frequency biological streams exhibit intense temporal autocorrelation, which can cause standard models to overfit on specific timeline segments. By utilizing independent bootstrap samples for each tree and forcing a structural 80/20 chronological split inside isolated condition blocks, the Random Forest maintains strict out-of-sample operational testing validity [WESAD].


## Analysis and Results

### Data Exploration and Visualization

-   Random forest script after normalization



A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{python}
import pandas as pd
from sklearn.ensemble import RandomForestClassifier
from sklearn.metrics import classification_report, f1_score, precision_score, recall_score

```

```{r, warning=FALSE, echo=TRUE}
# Load Data
print("=== Starting Random Forest Model Loop ===")

# 1. READ PRE-SPLIT, PRE-NORMALIZED FILES
print("Loading data splits...")
train_df = pd.read_csv("wesad_wrist_train.csv")
test_df = pd.read_csv("wesad_wrist_test.csv")

# 2. FEATURE ENGINEERING: COMPUTE TEMPORAL LAG ARRAYS WITH ECG INCLUDED
# Updated default list to scan for ECG, EDA, and TEMP features dynamically
def apply_time_lags(df, features=['ECG', 'EDA', 'TEMP']):
    # Filter features list to match only columns that exist inside the CSV file
    valid_features = [f for f in features if f in df.columns]
    print(f" -> Engineering lag features for existing channels: {valid_features}")
    
    lagged_groups = []
    # Loop ensures lags are calculated within each subject timeline independently
    for subject, group in df.groupby('subject'):
        group_copy = group.sort_index().copy()
        for col in valid_features:
            for lag in range(1, 5): # 4 lags = 1 second of context at 4Hz
                group_copy[f'{col}_lag_{lag}'] = group_copy[col].shift(lag)
        lagged_groups.append(group_copy.dropna())
    return pd.concat(lagged_groups, ignore_index=True)

print("Engineering lag context windows...")
train_lagged = apply_time_lags(train_df)
test_lagged = apply_time_lags(test_df)

# Filter feature columns out from metadata columns
feature_columns = [c for c in train_lagged.columns if c not in ['subject', 'label']]

# ====================================================
# 3. INDEPENDENT SUBJECT TRAINING LOOP
# ====================================================
for subject, group in train_lagged.groupby('subject'):
    print(f"\nTraining Random Forest Classifier for {subject}...")
    
    # Isolate training inputs for this subject
    X_train = group[feature_columns]
    y_train = group['label']
    
    # Isolate testing inputs from the matching hidden test split file
    test_group = test_lagged[test_lagged['subject'] == subject]
    X_test = test_group[feature_columns]
    y_test = test_group['label']
    
    # Initialize Random Forest with balanced class weights to address label skew
    rf_model = RandomForestClassifier(n_estimators=100, class_weight='balanced', random_state=42)
    rf_model.fit(X_train, y_train)
    
    # Evaluate performance
    y_pred = rf_model.predict(X_test)
    
    macro_f1 = f1_score(y_test, y_pred, average='macro')
    macro_prec = precision_score(y_test, y_pred, average='macro', zero_division=0)
    macro_rec = recall_score(y_test, y_pred, average='macro', zero_division=0)
    
    print(f"[{subject} Metrics Summary]")
    print(f" -> Macro Precision : {macro_prec:.4f}")
    print(f" -> Macro Recall    : {macro_rec:.4f}")
    print(f" -> Macro F1-Score  : {macro_f1:.4f}")
    print("\nDetailed Per-Class Performance:")
    print(classification_report(y_test, y_pred, labels=[1, 2, 3], target_names=['Neutral', 'Stress', 'Amusement'], zero_division=0))

print("\n=== Random Forest Execution Loop Complete ===")
 

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
Oyeleye M, Chen T, Titarenko S, Antoniou G. A Predictive Analysis of Heart Rates Using Machine Learning Techniques. Int J Environ Res Public Health. 2022 Feb 19;19(4):2417. doi: 10.3390/ijerph19042417. PMID: 35206603; PMCID: PMC8872524. 

Garg P, Santhosh J, Dengel A, Ishimaru S. Stress Detection by Machine Learning and Wearable Sensors. IUI '21 Companion. 2021. doi: 10.1145/3397482.3450732

Schmidt P, Reiss A. WESAD (Wearable Stress and Affect Detection). ACM ICMI 2018. doi: 10.1145/3242969.3242985

Kontopoulou VI, Panagopoulos AD, Kakkos I, Matsopoulos GK. A Review of ARIMA vs. Machine Learning Approaches for Time Series Forecasting in Data Driven Networks. Future Internet. 2023;15(8):255. doi: 10.3390/fi15080255

Ziyadidegan S, Sadeghi N, Razavi M, Baharlouei E, Janfaza V, Kazeminasab S, Pesarakli H, Javid AH, Sasangohar F. Quantifying Mental Stress Using Cardiovascular Responses: A Scoping Review. 2024. PMC12300869.