---
title: "Stress State Prediction from Wearable Sensors: A Comparative Study of ARIMA and Random Forest on the WESAD Dataset"
subtitle: "This is a Report Template Quarto"
author: "Indupriya Uppaluri, Virginia Vaughan, Chawan Sreelekha (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)



## Introduction

### Paper 1: Stress Detection by Machine Learning and Wearable Sensors

Garg et al. (2021) used WESAD dataset to compare several models including k-NN, Linear Discriminant Analysis, Random Forest, AdaBoost, and Support Vector Machine for both binary (stress vs. non-stress) and three-class (neutral vs. stress vs. amusement) classification which is a Machine Learning technique. During Garg et al. study Random forest's score was high enough to choose the model.

### Paper 2: Introducing WESAD, a Multimodal Dataset for Wearable Stress and Affect Detection

Schmidt et al. (2018) proposed WESAD dataset which was collected from 15 subjects using wrist-worn and chest-worn devices, capturing physiological signals including electrocardiogram (ECG), electrodermal activity (EDA), electromyogram (EMG), respiration, body temperature, blood volume pulse, and three-axis acceleration. This paper is foundational to our project as it describes the exact dataset we are using and establishes the key physiological signals most relevant to stress detection.


Paper 7 :A Review of ARIMA vs. Machine Learning for Time Series Forecasting
Kontopoulou et al. (2023) research paper compares ARIMA and Machine Learning algorithms for time-series forecasting problems. The research suggests that ARIMA model is broadly used since its easy to understand and apply, but authors state that the swift to Machine learning models occurred without assessing ARIMA in all conceptual frameworks. The paper summarizes that ARIMA is more strategic choice in structured, minimalistic datasets, whereas machine learning dominates with large and complex datasets.
Paper 8: Quantifying Mental Stress Using Cardiovascular Responses — Scoping Review 

Ziyadidegan et al. (2025) research paper studied how cardiovascular responses, particularly heart rate (HR) and heart rate variability (HRV), are used to quantify mental stress. The authors found that heart rate activity consistently changes during stressful events with HR increasing while HRV measurements such as RMSSD and SDNN decrease, confirming heart activity as a reliable stress indicator.  The majority of reviewed studies used descriptive statistical methods including t-tests, ANOVA, and correlation analysis to assess differences between stress and non-stress conditions. Importantly, the review highlights time series techniques including ARIMA, detrended fluctuation analysis, and autocorrelation plots as valuable tools for studying heart rate rhythms and temporal patterns associated with mental stress. Although ARIMA appeared in only 3 of the 119 reviewed studies, the authors emphasize its potential as an emerging and underutilized method for capturing temporal changes in physiological stress responses. Most studies were conducted in controlled laboratory settings comparable to the WESAD dataset collection protocol, and this paper is directly relevant to our project as it provides methodological justification for applying ARIMA to physiological time series signals for stress state detection.



## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{python}
import pandas as pd
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
