---
title: "Stress State Prediction from Wearable Sensors: A Comparative Study of ARIMA and Random Forest on the WESAD Dataset"

author: "Indupriya Uppaluri, Virginia Vaughan, Chawan Sreelekha (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)



## Introduction

Stress detection in real-time, and more importantly biomarkers, have gained interest as both military and corporate facilities aim to improve the quality of worker health and performance. This has let to extensive research using commercial wearables that record signals such as beats per minute (BPM), heartrate (HR), heart rate variability (HRV), galvanic skin conductance (EDA), respiration (RESP), and temperature (TEMP). These physiological signals are used in tandem with both stochastic models such as ARIMA and machine learning such as Random Forest. Both types of models offer their own positives and negatives when predicting physiological data. For example, ARIMA is a more traditional, linear-based model, requires less data for training and extrapolation, while Random Forest is more data-intensive but does not assume linearity, making it better suited for modeling complex patterns. 

In this paper, we hope to explore whether it is possible for the simpler ARIMA model to predict stress state at the same accuracy of a machine learning model like Random Forest. If the ARIMA can perform similar to or greater, this makes an argument that it is still possible to predict stress in physiological signals without the intensive training needed in machine learning, despite ARIMA’s assumptions and limitations. To explore this, we chose the WESAD dataset, which contains multimodal physiological data collected from participants as they underwent different activities, such as no-stress, stress, and amusement. Based on our literature review, we have found that the best physiological signals for predicting stress state that are also provided by the dataset include HR, HRV, and EDA, and that these metrics can be enhanced with secondary metrics such as RESP and TEMP. One paper, [@oyeleye2022] found that the accuracy of ARIMA prediction can be heavily improved using the “rolling window” approach. This helped the model predict based on the signal during different windows of time instead of treating each sample independently of the other. In our methodology we plan to create a pipeline to preprocess the data, compute our feature metrics, derive rolling windows, and predict a given metric using both ARIMA and Random Forest in order to compare the model’s accuracy. 


ARIMA, which stands for Autoregressive Integrated Moving Average, is a classical statistical model widely used for time-series forecasting that captures temporal dependencies within a signal using its own past values and error terms. [@kontopoulou2023] found that ARIMA remains a broadly used and interpretable approach, particularly well-suited for structured, lower-dimensional datasets where its linearity assumptions hold reasonably well. Critically, [@ziyadidegan2025] identified ARIMA as an underutilized but promising method specifically for physiological stress research, appearing in only 3 of 119 reviewed cardiovascular stress studies despite its recognized capacity to capture temporal rhythms in cardiovascular and electrodermal signals. Given that ECG-derived metrics reflect the electrical activity of the heart and EDA captures the skin's electrodermal response to psychological arousal, both of which change measurably during stress, the temporal structure of these signals makes them theoretically well-suited for time-series modeling through ARIMA with a rolling window approach. If stress-related changes in ECG and EDA follow predictable temporal patterns, ARIMA may be capable of detecting stress states without the computational demands of machine learning, a question this study directly investigates. 

A systematic review of the literature on the classification of multimodal sympathetic and parasympathetic nervous system arousal showed that EDA, a measure of sympathetic nervous system activity, and HRV, a measure of parasympathetic nervous system activity, are a more complete reconstruction of autonomic state than either signal alone. We therefore chose respiration as our indicator of stress because it provides an additional measurement of the changes in breathing pattern that occur when a person is stressed, as well as the electrodermal response measured by EDA.

Random Forest is an ensemble machine learning method that builds multiple decision trees and aggregates their outputs to produce robust classifications, making no assumptions about the linearity or distribution of the underlying data. [@garg2021] evaluated Random Forest alongside several other classifiers including k-NN, Linear Discriminant Analysis, AdaBoost, and Support Vector Machine on the WESAD dataset, finding that Random Forest achieved consistently strong performance across both binary stress versus non-stress and three-class neutral, stress, and amusement classification tasks. [@schmidt2018], who originally introduced the WESAD dataset, a multimodal physiological signal collection from 15 subjects recorded across controlled stress, baseline, and amusement conditions using wrist-worn and chest-worn devices capturing signals including ECG and EDA among others, also conducted preliminary Random Forest experiments achieving 88.33% baseline accuracy, further establishing it as the benchmark comparator for this study. Together, these findings position Random Forest as a well-validated and high-performing method for stress classification on WESAD, and the central question of this study is whether the simpler, interpretable ARIMA model can achieve comparable predictive performance on ECG and EDA signals alone.


## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{python}
import pandas as pd
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
Oyeleye M, Chen T, Titarenko S, Antoniou G. A Predictive Analysis of Heart Rates Using Machine Learning Techniques. Int J Environ Res Public Health. 2022 Feb 19;19(4):2417. doi: 10.3390/ijerph19042417. PMID: 35206603; PMCID: PMC8872524. 

Garg P, Santhosh J, Dengel A, Ishimaru S. Stress Detection by Machine Learning and Wearable Sensors. IUI '21 Companion. 2021. doi: 10.1145/3397482.3450732

Schmidt P, Reiss A. WESAD (Wearable Stress and Affect Detection). ACM ICMI 2018. doi: 10.1145/3242969.3242985

Kontopoulou VI, Panagopoulos AD, Kakkos I, Matsopoulos GK. A Review of ARIMA vs. Machine Learning Approaches for Time Series Forecasting in Data Driven Networks. Future Internet. 2023;15(8):255. doi: 10.3390/fi15080255

Ziyadidegan S, Sadeghi N, Razavi M, Baharlouei E, Janfaza V, Kazeminasab S, Pesarakli H, Javid AH, Sasangohar F. Quantifying Mental Stress Using Cardiovascular Responses: A Scoping Review. 2024. PMC12300869.